Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-08 src/ms.tex
Initial commit

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\usepackage{xspace}
\usepackage{amsmath}
\usepackage{longtable}

@@

\newcommand{\rocrit}{$\mathrm{Ro_{crit}}$\xspace}

\newcommand{\kms}{km~s$^{-1}$}
\newcommand{\ms}{m~s$^{-1}$}
\newcommand{\gcc}{g~cm$^{-3}$}
\newcommand{\masyr}{mas~yr$^{-1}$}
\newcommand{\err}{\textit{$\pm$}}

\newcommand{\Mstar}{\ensuremath{M_{\star}}\xspace}
\newcommand{\Rstar}{\ensuremath{R_{\star}}\xspace} 
\newcommand{\Lstar}{\ensuremath{L_{\star}}\xspace} 
\newcommand{\teff}{\ensuremath{T_{\mathrm{eff}}}\xspace}  
\newcommand{\logg}{\ensuremath{\log g}\xspace} 
\newcommand{\feh}{[Fe/H]\xspace}
\newcommand{\vsini}{\ensuremath{v \sin i}\xspace} 
\newcommand{\prot}{\ensuremath{P_\mathrm{rot}}\xspace}
\newcommand{\rvar}{\ensuremath{R_\mathrm{var}}\xspace}
\newcommand{\vtan}{\ensuremath{v_\mathrm{tan}}\xspace}
\newcommand{\vb}{\ensuremath{v_\mathrm{b}}\xspace}

\newcommand{\mearth}{$M_\oplus$\xspace}
\newcommand{\rearth}{$R_\oplus$\xspace}

\newcommand{\msun}{$M_\odot$\xspace}
\newcommand{\rsun}{$R_\odot$\xspace}
\newcommand{\lsun}{$L_\odot$\xspace}
\newcommand{\rhosun}{$\rho_\odot$\xspace}
\newcommand{\mstar}{$M_*$\xspace}
\newcommand{\rstar}{$R_*$\xspace}
\newcommand{\lstar}{$L_*$\xspace}
\newcommand{\rhostar}{$\rho_*$\xspace}

\newcommand{\logage}{$\text{log(age)}$\xspace}
\newcommand{\rp}{$R_P$\xspace}
\newcommand{\rprstar}{$R_P/R_\star$\xspace}

\newcommand{\logrp}{$\log_{10}(R_P/R_\oplus)$\xspace}
\newcommand{\deltaage}{$\Delta \log_{10}(\mathrm{age~yr}^{-1})$\xspace}
\newcommand{\deltamstar}{$\Delta M_*$\xspace}
\newcommand{\deltateff}{$\Delta T_\mathrm{eff}$\xspace}

\newcommand{\mjup}{$M_\mathrm{Jup}$\xspace}
\newcommand{\galex}{\textit{GALEX}\xspace}
\newcommand{\gaia}{\textit{Gaia}\xspace}
\newcommand{\kepler}{\textit{Kepler}\xspace}
\newcommand{\spitzer}{\textit{Spitzer}}
\newcommand{\hipparcos}{\textit{Hipparcos}}
\newcommand{\ktwo}{\textit{K2}\xspace}
\newcommand{\tess}{\textit{TESS}\xspace}
\newcommand{\emcee}{\texttt{emcee}\xspace}
\newcommand{\stardate}{\texttt{stardate}\xspace}
\newcommand{\python}{\textsc{python}\xspace}

\newcommand{\ymg}{[Y/Mg]\xspace}
\newcommand{\yal}{[Y/Al]\xspace}
\newcommand{\alphafe}{[$\alpha$/Fe]\xspace}
\newcommand{\dexgyr}{dex~Gyr$^{-1}$\xspace}

@@

% \title{The Rossby Ridge: a Pile-up of Rapidly Rotating Main-Sequence Stars with a Broad Range of Ages}
\title{A Pile-up of Rapidly Rotating Main-Sequence Stars with a Broad Range of Ages: \\Further Evidence for Weakened Magnetic Braking}

@@

\author[0000-0001-6534-6246]{Trevor J.\ David}
\affil{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}
\affil{Department of Astrophysics, American Museum of Natural History, Central Park West at 79th Street, New York, NY 10024, USA}

\author[0000-0003-4540-5661]{Ruth Angus}
\affil{Department of Astrophysics, American Museum of Natural History, Central Park West at 79th Street, New York, NY 10024, USA}
\affil{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}
\affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY, USA}

\author[0000-0002-2792-134X]{Jason L.~Curtis}
\affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY, USA}

\author[0000-0002-4284-8638]{Jennifer van Saders}
\affil{Institute for Astronomy, University of Hawai’i, Honolulu, HI, USA}

\author[0000-0002-3011-4784]{Gabriella Contardo}
\affil{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}

\author{Other authors TBD}

%Lucy Lu
%Joel Zinn
%Adrian Price-Whelan

\begin{abstract}
We combine stellar surface rotation periods determined from NASA's Kepler mission with spectroscopic parameters derived from the California-Kepler Survey and LAMOST--Kepler survey to demonstrate the existence of a long-period pile-up in the effective temperature--rotation period distribution of main-sequence stars with temperatures exceeding 5800~K. Stars in the long-period pile-up are found to have a wide range of ages ($\sim$2--6~Gyr), meaning that, along the ridge, rotation period is strongly predictive of a star's surface temperature but only weakly predictive of its age. In the effective temperature range of 5000--6250~K the long-period pile-up is well-described by a single-parameter model of constant Rossby number, with a critical value of $\mathrm{Ro_{crit}} \sim$\textbf{XYZ}. The long-period pile-up was previously predicted by \citet{vanSaders2019} as a consequence of weakened magnetic braking, in which stellar spin-down stalls once stars reach a critical Rossby number. These results imply that surface rotation periods have limited predictive power for the age determination of F- and early G-type stars older than $\sim$1--2~Gyr. \textbf{We additionally detect a secondary pile-up associated with the short-period edge of the rotation period distribution.}
\end{abstract}

\keywords{Stellar rotation (1629) --- Solar analogs (1941) --- Stellar evolution (1599) --- Stellar magnetic fields (1610) --- Stellar winds(1636)}

@@

Here we examine the rotation period distribution of main-sequence stars observed by Kepler, leveraging the recent release of precise spectroscopic parameters from large-scale surveys, to demonstrate the existence of a long-period pile-up for F- and early G-type stars. \textbf{In this work we refer to this feature as the long-period pile-up, the Rossby ridge, or simply the ridge.} We argue that several aspects of this pile-up feature are consistent with expectations from the WMB model. We discuss our sample in \S\ref{sec:sample}, describe the steps of our analysis in \S\ref{sec:analysis}, and present our conclusions in \S\ref{sec:conclusions}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{surveys.pdf}
    \caption{The \teff-\prot plane using rotation periods from \citet{McQuillan2014} or, in the case of the CKS sample, \citet{McQuillan2013}, with \teff originating from the source denoted at top. The \citet{McQuillan2014} \teff values originate from the Kepler Input Catalog \citep[KIC,][]{Brown2011} or \citet{Dressing2013} for low-mass stars. The orange point in each panel indicates the Sun's equatorial rotation period, with the errorbar capturing the range of periods measured from the differentially rotating surface.}
    \label{fig:surveys}
\end{figure*}

\subsection{California--Kepler Survey} \label{sec:cks}
The CKS project gathered high-resolution spectroscopy for 1305 Kepler planet host stars \citep{Petigura2017}. CKS spectra were acquired with the Keck/HIRES spectrograph \citep{Vogt1994} and spectroscopic parameters were determined by averaging parameters from the SpecMatch pipeline \citep{Petigura2015} and SME@XSEDE, a Python implementation of the Spectroscopy Made Easy pipeline \citep{Valenti1996}. The internal errors on \teff from the CKS catalog are 60~K.

We compiled rotation periods for these stars from a variety of literature sources including \citet{McQuillan2013, Mazeh2015} and \citet{Angus2018}. For each star in the sample we then visually inspected the Kepler light curve folded on all available literature periods, as well as the first harmonics and sub-harmonics of those periods, and recorded our preferred period along with a reliability flag. Our procedure is explained in detail in \S2.1 of \citet{David2021}, and rotation period vetting sheets for each Kepler Object of Interest (KOI) are publicly available through Zenodo.\footnote{\url{http://10.0.20.161/zenodo.4645437}} We note that the vast majority of stars in the CKS sample host small planets ($R_P < 4$~\rearth) and as such it is not expected that the host stars have experienced tidal spin-up from the planets.

In addition to the original CKS catalog, we also cross-matched our sample with the catalogs of \citet{Brewer2018} and \citet{Martinez2019}, both of which presented spectroscopic parameters for CKS stars based on independent analysis of the same spectra. The \citet{Brewer2018} study, referred to here as SPOCS, also published elemental abundances and ages from isochrone fitting for the CKS sample.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{ridge.pdf}
    \caption{The \teff-\prot plane for the CKS sample. Point colors are scaled to the CKS ages determined from isochrone fitting. The source of \prot is denoted above each panel, where \citet{David2021} is a compilation of vetted periods, rather than a source of original measurements. The black trapezoid indicates the approximate area of the ridge. The grey curves indicate empirical cluster sequences from \citet{Curtis2020}, corresponding to ages of $\sim$2.7, 1, 0.67, and 0.12~Gyr from top to bottom.}
    \label{fig:ridge}
\end{figure*}

\subsection{LAMOST--Kepler} \label{sec:lamost}
The LAMOST--Kepler project derived homogeneous spectroscopic parameters from low-resolution ($R\sim$~1800) LAMOST DR5 spectra for approximately 40\% of the Kepler field \citep{Zong2018, Xiang2019}. We cross-matched the LAMOST--Kepler stellar parameter catalog of \citet{Xiang2019} with the \citet{McQuillan2014} rotation period catalog, which published rotation periods for $>$34000 Kepler targets. The \citet{Xiang2019} catalog derived stellar parameters using the DD-Payne pipeline, which builds on the method of \citet{Ting2017b} by incorporating elements of the Cannon \citep{Ness2015}.

We matched 10844 LAMOST targets to 10650 unique Kepler IDs. For the Kepler sources with duplicate cross-matched LAMOST sources we kept the source with a brighter Gaia DR2 $G$ magnitude. We additionally selected sources with 3000~K$< \teff <$~8000~K, 3$<$~\logg~$<$5 dex, and -2~$<$[Fe/H]$<$~2 dex. The final sample contains 10516 unique targets with well-determined \prot and \teff (having a median error of 24~K). There is negligible overlap (3 stars) between our LAMOST--Kepler samples and the CKS samples since the \citet{McQuillan2014} did not publish rotation periods for KOIs, which were the targets of the CKS project.

\textbf{We additionally cross-matched the LAMOST--Kepler catalog with the CKS catalog...Note, LAMOST catalog contains duplicates from multiple visits. Describe how we handle these.}

\subsection{APOGEE--Kepler}
The Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE,][]{Majewski2017} is a large-scale, high-resolution ($R \sim 22500$) stellar spectroscopic survey conducted at $H$-band as part of the Sloan Digital Sky Survey \citep[SDSS-IV][]{Blanton2017}. The spectroscopic analysis pipeline for SDSS DR16 is described in \citet{Jonsson2020}. We used Gaia DR2 source IDs \citep{Gaia2016, Gaia2018} to cross-match Megan Bedell's Gaia--Kepler catalog\footnote{\url{https://gaia-kepler.fun/}} with the APOGEE DR16 catalog \citep{Ahumada2020}. Kepler IDs were then used to cross-match this table with the \citet{McQuillan2014} catalog. While the current overlap between Kepler targets and APOGEE is small compared to the LAMOST catalog, APOGEE DR17 will contain more dwarf stars and provide a better resource for studies such as ours. The focus of this work are overdensities in the \teff--\prot plane, and as these appear to be less prominent when using  APOGEE DR16 temperatures (Figure~\ref{fig:surveys}) we conclude that LAMOST and CKS provide more precise estimates of \teff and do not analyze the APOGEE sample further.

\subsection{The Sun}
To place the Sun in the context of the long-period pile-up, we use up-to-date estimates of the Sun's effective temperature, rotation period, and age. Following the IAU 2015 Resolution B3 \citep{Mamajek2015}, we take the nominal effective temperature of the Sun to be $\mathcal{T}^\mathrm{N}_\mathrm{eff,\odot} = 5772$~K. For the solar rotation period we adopt the equatorial rotation period of $P_\mathrm{eq,\odot}~\approx~25$~d with an upper error that encompasses the period of $\approx$~36~d measured near the poles from differential rotation studies \citep[][and references therein]{Thompson2003}.  The age of the Sun is assumed to be 4.567~Gyr from Pb-Pb dating of calcium-aluminum inclusions and chondrules recovered from primitive meteorites \citep{Bahcall1995}.

%\prot = 26.09~d, with minimum and maximum periods of $P_\mathrm{min}=24.5$ and $P_\mathrm{max}=28.5$, from \citet{Donahue1996}. Measurements of differential rotation on the solar surface have found an equatorial period of approximately 25 days, increasing to about 36 days at the poles \citep[][and references therein]{Thompson2003}.

@@

\subsection{Initial observations}
A pile-up of stars along the long-period edge for stars with \teff~$>5800$~K was first noted when examining the \teff-\prot plane for the CKS sample, using the CKS \teff values and the vetted rotation period compilation from \citet{David2021}. Sourcing rotation periods from \citet{McQuillan2013, Mazeh2015} and \citet{Angus2018} revealed that this pile-up is still apparent when adopting periods uniformly from a single catalog (Figure~\ref{fig:ridge}).

The location of the pile-up approximately corresponds with an empirical hybrid cluster sequence derived by \citet{Curtis2020} from members of the NGC~6819 (age~$\sim$~2.5~Gyr) and Ruprecht~147 (age~$\sim$~2.7~Gyr) open clusters (Figure~\ref{fig:ridge}). As we show in \S\ref{subsec:ages}, stars along the pile-up have a range of ages. Therefore, the observation that the edge approximately corresponds with the $\sim$2.5--2.7~Gyr cluster sequence implies that stars with \teff$\lesssim 5800~K$ have already piled up onto the edge by or before this timescale. Similarly, the edge is well-separated from the empirical $\sim$1~Gyr cluster sequence based on rotation rates in the NGC~6811 cluster \citep{Curtis2020}, implying that it takes F-type stars $>1$~Gyr to converge onto the long-period edge. These observations are in accordance with predictions from the WMB model which suggest the pile-up forms on a timescale of 2--3~Gyr \citep{vanSaders2019}.

In both the CKS and LAMOST samples, there is a change of slope along the ridge, with an inflection point corresponding closely to the Kraft break at $\approx$~6250~K, the point at which convective envelopes become vanishingly thin \citep{Kraft1967}. A piecewise linear fit to the ridge in the CKS sample confirmed that the inflection point occurs at \teff~=~6250~$\pm$32~K, where the uncertainty was estimated from Markov chain Monte Carlo (MCMC) sampling. \textbf{Some interpretation.}

>>>>

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{longtable}

\newcommand{\taucz}{$\tau_\mathrm{cz}$\xspace}
\newcommand{\rocrit}{$\mathrm{Ro_{crit}}$\xspace}
\newcommand{\rocritfinal}{$\mathrm{Ro_{crit}}~\lesssim~2$\xspace}
\newcommand{\lamostkep}{LAMOST--Kepler\xspace}
\newcommand{\lamostmcq}{LAMOST--McQuillan\xspace}
\newcommand{\teffmin}{5800~K\xspace}
\newcommand{\agerange}{$\sim$2--6~Gyr\xspace}
\newcommand{\agemin}{$\sim$2--3~Gyr\xspace}

@@

\newcommand{\Mstar}{\ensuremath{M_{\star}}\xspace}
\newcommand{\Rstar}{\ensuremath{R_{\star}}\xspace} 
\newcommand{\Lstar}{\ensuremath{L_{\star}}\xspace} 
\newcommand{\teff}{\ensuremath{T_{\mathrm{eff}}}\xspace}  
\newcommand{\logg}{\ensuremath{\log g}\xspace} 
\newcommand{\feh}{[Fe/H]\xspace}
\newcommand{\vsini}{\ensuremath{v \sin i}\xspace} 
\newcommand{\msun}{$M_\odot$\xspace}
\newcommand{\rsun}{$R_\odot$\xspace}
\newcommand{\lsun}{$L_\odot$\xspace}
\newcommand{\tsun}{$T_\mathrm{eff,\odot}$\xspace}
\newcommand{\rhosun}{$\rho_\odot$\xspace}
\newcommand{\mstar}{$M_*$\xspace}
\newcommand{\rstar}{$R_*$\xspace}
\newcommand{\lstar}{$L_*$\xspace}
\newcommand{\rhostar}{$\rho_*$\xspace}
\newcommand{\logage}{$\text{log(age)}$\xspace}
\newcommand{\mearth}{$M_\oplus$\xspace}
\newcommand{\rearth}{$R_\oplus$\xspace}
\newcommand{\prot}{\ensuremath{P_\mathrm{rot}}\xspace}
\newcommand{\rvar}{\ensuremath{R_\mathrm{var}}\xspace}
\newcommand{\rper}{\ensuremath{R_\mathrm{per}}\xspace}

\newcommand{\gaia}{\textit{Gaia}\xspace}
\newcommand{\kepler}{\textit{Kepler}\xspace}
\newcommand{\spitzer}{\textit{Spitzer}}
\newcommand{\tess}{\textit{TESS}\xspace}
\newcommand{\emcee}{\texttt{emcee}\xspace}
\newcommand{\stardate}{\texttt{stardate}\xspace}
\newcommand{\python}{\textsc{python}\xspace}

\newcommand{\cca}{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}
\newcommand{\amnh}{Department of Astrophysics, American Museum of Natural History, Central Park West at 79th Street, New York, NY 10024, USA}
\newcommand{\columbia}{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY, USA}
\newcommand{\ifa}{Institute for Astronomy, University of Hawai’i, Honolulu, HI, USA}

@@

% \title{The Rossby Ridge: a Pile-up of Rapidly Rotating Main-Sequence Stars with a Broad Range of Ages}
%\title{A Pile-up of Rapidly Rotating Main-Sequence Stars with a Broad Range of Ages: \\Further Evidence for Weakened Magnetic Braking}
%\title{Two Pile-ups in the Temperature--Period Distribution of Main-Sequence Stars:\\Possible Evidence for Weakened Magnetic Braking and Core-Envelope Coupling}
\title{Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution}

@@

\author[0000-0001-6534-6246]{Trevor J.\ David}
\affil{\cca}
\affil{\amnh}

\author[0000-0003-4540-5661]{Ruth Angus}
\affil{\amnh}
\affil{\cca}
\affil{\columbia}

\author[0000-0002-2792-134X]{Jason L.~Curtis}
\affil{\columbia}

\author[0000-0002-4284-8638]{Jennifer L.~van Saders}
\affil{\ifa}

\author[0000-0002-3011-4784]{Gabriella Contardo}
\affil{\cca}

\author[0000-0003-4769-3273]{Yuxi Lu}
\affil{\columbia}
\affil{\amnh}

\author[0000-0002-7550-7151]{Joel C.~Zinn}
\affil{\amnh}

\begin{abstract}
We combine stellar surface rotation periods determined from NASA's Kepler mission with spectroscopic parameters from the California-Kepler Survey and LAMOST--Kepler survey to demonstrate the existence of a pile-up at the long-period edge of the effective temperature--rotation period distribution of main-sequence stars with temperatures exceeding $\sim$\teffmin. Stars in the long-period pile-up are found to have a wide range of ages (\agerange), meaning that, along the pile-up, rotation period is strongly predictive of a star's surface temperature but only weakly predictive of its age. In the effective temperature range of 5000--6250~K the long-period edge is well-described by a curve of constant Rossby number, with a critical value of \rocritfinal. The long-period pile-up was previously predicted by \citet{vanSaders2019} as a consequence of weakened magnetic braking, in which wind-driven angular momentum losses cease once stars reach a critical Rossby number. These results imply that surface rotation periods have limited predictive power for the age determination of F- and early G-type stars older than \agemin. Using kernel density estimation, we additionally detect a secondary pile-up associated with the short-period edge of the rotation period distribution. {\color{blue}The short-period pile-up is not a prediction of the weakened magnetic braking hypothesis, but may instead be related to a phase of delayed spin-down due to core-envelope coupling previously proposed by \citet{Curtis2020} to explain the overlapping rotation sequences of low-mass members of differently aged open clusters. Together, the short- and long-period pile-ups create an underdensity of intermediate rotation periods which aligns with the period gap initially discovered by \citet{McQuillan2013b} in M-type stars.}
\end{abstract}

\keywords{Stellar rotation (1629) --- Solar analogs (1941) --- Stellar evolution (1599) --- Stellar magnetic fields (1610) --- Stellar winds (1636)}

@@

Here we examine the rotation period distribution of main-sequence stars observed by Kepler, leveraging the recent release of precise spectroscopic parameters from large-scale surveys, to demonstrate the existence of pile-ups at the long- and short-period edges of the \teff--\prot distribution of solar-type stars. We discuss our sample in \S\ref{sec:sample}, describe the steps of our analysis in \S\ref{sec:analysis}, discuss some implications of these results in \S\ref{sec:discussion}, and present our conclusions in \S\ref{sec:conclusions}.

%We argue that several aspects of the long-period pile-up feature are consistent with expectations from the WMB model. The short-period pile-up, by comparison, may be due a temporary epoch of stalled wind-driven losses due to core-envelope coupling.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{./figures/surveys.pdf}
    \caption{The \teff-\prot plane using rotation periods from \citet{McQuillan2014} or, in the case of the CKS sample, \citet{McQuillan2013}, with \teff originating from the source denoted at top. The \citet{McQuillan2014} \teff values originate from the Kepler Input Catalog \citep[KIC,][]{Brown2011} or \citet{Dressing2013} for low-mass stars. The orange point in each panel indicates the Sun's equatorial rotation period, with the errorbar capturing the range of periods measured from the differentially rotating surface.}
    \label{fig:surveys}
\end{figure*}

\subsection{California--Kepler Survey} \label{sec:cks}
The CKS project gathered high-resolution spectroscopy for 1305 Kepler planet host stars \citep{Petigura2017}. CKS spectra were acquired with the Keck/HIRES spectrograph \citep{Vogt1994} and spectroscopic parameters were determined by averaging parameters from the SpecMatch pipeline \citep{Petigura2015} and SME@XSEDE, a Python implementation of the Spectroscopy Made Easy pipeline \citep{Valenti1996}. The internal errors on \teff from the CKS catalog are 60~K. The metallicity distribution of the CKS sample is centered near solar, with a mean and standard deviation of +0.03~dex and 0.18~dex, respectively.

We compiled rotation periods for these stars from a variety of literature sources including \citet{McQuillan2013, Mazeh2015} and \citet{Angus2018}. For each star in the sample we then visually inspected the Kepler light curve folded on all available literature periods, as well as the first harmonics and sub-harmonics of those periods, and recorded our preferred period along with a reliability flag. Our procedure is explained in detail in \S2.1 of \citet{David2021}, and rotation period vetting sheets for each Kepler Object of Interest (KOI) are publicly available through Zenodo.\footnote{\url{http://10.0.20.161/zenodo.4645437}} The vast majority of stars in the CKS sample host small planets ($R_P < 4$~\rearth) and as such it is not expected that the host stars have experienced tidal spin-up from the planets.

In addition to the original CKS catalog, we also cross-matched our sample with the catalogs of \citet{Brewer2018} and \citet{Martinez2019}, both of which presented spectroscopic parameters for CKS stars based on independent analysis of the same spectra. The \citet{Brewer2018} study, referred to here as SPOCS, also published elemental abundances and ages from isochrone fitting for the CKS sample. We additionally cross-matched the CKS catalog with the LAMOST--Kepler catalog \citep{Xiang2019} which is described further in \S\ref{subsec:lamost}. We compare the \teff--\prot distributions of the CKS sample using \teff and rotation periods from a variety of sources in Appendix~\ref{app:teffprot}.

\subsection{LAMOST--Kepler} \label{subsec:lamost}
The LAMOST--Kepler project derived homogeneous spectroscopic parameters from low-resolution ($R\sim$~1800) LAMOST DR5 spectra for approximately 40\% of the Kepler field \citep{Zong2018, Xiang2019}. We cross-matched the LAMOST--Kepler stellar parameter catalog of \citet{Xiang2019} with the \citet{McQuillan2014} rotation period catalog, which published rotation periods for $>$34000 Kepler targets. The \citet{Xiang2019} catalog derived stellar parameters using the DD-Payne pipeline, which builds on the method of \citet{Ting2017b} by incorporating elements of the Cannon \citep{Ness2015}.

We matched 10844 LAMOST targets to 10650 unique Kepler IDs. For the Kepler sources with duplicate cross-matched LAMOST sources we kept the source with a brighter Gaia DR2 $G$ magnitude. We additionally selected sources with 3000~K~$<~\teff~<$~8000~K, 3~$<$~\logg~$<$~5~dex, and -2~$<$~[Fe/H]~$<$~2~dex. The final sample contains 10516 unique targets with well-determined \prot and \teff (having a median error of 24~K). The metallicity distribution of the LAMOST--McQuillan sample is centered near solar, with a mean and standard deviation of -0.1~dex and 0.26~dex, respectively. There is negligible overlap (3 stars) between our LAMOST--McQuillan sample and the CKS sample since the \citet{McQuillan2014} did not publish rotation periods for KOIs, which were the targets of the CKS project.

\subsection{APOGEE--Kepler}
The Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE;][]{Majewski2017} is a large-scale, high-resolution ($R \sim 22500$) stellar spectroscopic survey conducted at $H$-band as part of the Sloan Digital Sky Survey \citep[SDSS-IV;][]{Blanton2017}. The spectroscopic analysis pipeline for SDSS DR16 is described in \citet{Jonsson2020}. We used Gaia DR2 source IDs \citep{Gaia2016, Gaia2018} to cross-match Megan Bedell's Gaia--Kepler catalog\footnote{\url{https://gaia-kepler.fun/}} with the APOGEE DR16 catalog \citep{Ahumada2020}. Kepler IDs were then used to cross-match this table with the \citet{McQuillan2014} catalog. While the current overlap between Kepler targets and APOGEE is small compared to the LAMOST catalog, APOGEE DR17 will contain more dwarf stars and provide a better resource for studies such as ours. The focus of this work are overdensities in the \teff--\prot plane, and as these appear to be less prominent when using  APOGEE DR16 temperatures (Figure~\ref{fig:surveys}) we conclude that LAMOST and CKS provide more precise estimates of \teff and do not analyze the APOGEE sample further.

\subsection{The Sun}
To place the Sun in the context of the long-period pile-up, we use up-to-date estimates of the Sun's effective temperature, rotation period, and age. Following the IAU 2015 Resolution B3 \citep{Mamajek2015}, we take the nominal effective temperature of the Sun to be $\mathcal{T}^\mathrm{N}_\mathrm{eff,\odot} = 5772$~K. For the solar rotation period we adopt the equatorial rotation period of $P_\mathrm{eq,\odot}~\approx~25$~d with an upper error that encompasses the period of $\approx$~36~d measured near the poles from differential rotation studies \citep[][and references therein]{Thompson2003}.  The age of the Sun is assumed to be 4.567~Gyr from Pb-Pb dating of calcium-aluminum inclusions and chondrules recovered from primitive meteorites \citep[][and references therein]{Bahcall1995}.

@@

\subsection{Initial observations}
A pile-up of stars along the long-period edge for stars with \teff~$>5800$~K was first noted when examining the \teff-\prot plane for the CKS sample, using the CKS \teff values and the vetted rotation period compilation from \citet{David2021}. Sourcing rotation periods from \citet{McQuillan2013, Mazeh2015} and \citet{Angus2018} revealed that this pile-up is still apparent when adopting periods uniformly from a single catalog (Figure~\ref{fig:ridge}). {\color{blue} A secondary pile-up at the short-period edge is also apparent, though less pronounced, in the \teff--\prot distribution of the \lamostmcq sample. As shown in Figure~\ref{fig:kde}, This secondary pile-up is seen most clearly through applying Gaussian kernel density estimation (KDE), which was performed with the \texttt{seaborn} package \citep{seaborn}.}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{./figures/kde.pdf}
    \caption{Gaussian kernel density estimation (blue contours) of the \teff--\prot distributions of the CKS--McQuillan, \lamostmcq, and asteroseismic \citet{Hall2021} samples, from left to right. Empirical cluster sequences from \citet{Curtis2020} are shown by the dark grey lines. The orange dashed lines show constant Rossby curves of fiducial values (see \S\ref{subsec:rossby}). The short-period pile-up can be observed in the LAMOST--McQuillan sample for \teff~$\gtrsim$~5500~K. The orange point indicates the Sun's temperature and equatorial rotation period, with the errorbar capturing the range of periods measured from its differentially rotating surface.}
    \label{fig:kde}
\end{figure*}

The location of the pile-up approximately corresponds with an empirical hybrid cluster sequence derived by \citet{Curtis2020} from members of the NGC~6819 (age~$\sim$~2.5~Gyr) and Ruprecht~147 (age~$\sim$~2.7~Gyr) open clusters (Figure~\ref{fig:kde}). As we show in \S\ref{subsec:ages}, stars along the pile-up have a range of ages. Therefore, the observation that the edge approximately corresponds with the $\sim$2.5--2.7~Gyr cluster sequence implies that stars with \teff$\lesssim 5800~K$ have already piled up onto the edge by or before this timescale. Similarly, the edge is well-separated from the empirical $\sim$1~Gyr cluster sequence based on rotation rates in the NGC~6811 cluster \citep{Curtis2020}, implying that it takes F-type stars $>1$~Gyr to converge onto the long-period edge. These observations are in accordance with predictions from the WMB model which suggest the pile-up forms on a timescale of 2--3~Gyr \citep{vanSaders2019}.
